---
title: "sm_predict"
author: "Daniel D. Sjoberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sm_predict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sm_predict()` function calculates kernel-smoothed predictions from regression models (i.e. outputs from the `predict()` function).  The following examples focus on time-to-event endpoints.

## Example 1 - Univariate

The function can be computationally intense on large data sets.  The following code was run to get the example object `sm_regression_ex`.

```
library(survival)
library(sjosmooth)
sm_predict_ex1 =
  sm_predict(
    data = cancertx,
    method = "coxph",
    formula = Surv(time) ~ marker,
    newdata = dplyr::data_frame(marker = seq(5, 15, 0.5), time = 1),
    type = "survival",
    verbose = TRUE
  )
```

The data in the `cancertx` data frame was simulated from a Cox regression model with linear predictor.  

$$
\textbf{X}\beta = 0.02 * age - 0.2 * marker
$$

In the model age and marker level are independent, and therefore, the univariate model is simply

$$
\textbf{X}\beta = 0.02 * age
$$


```{r remedy001}
library(sjosmooth)
library(survival)
library(dplyr)
library(ggplot2)
# loading data and example results from sjosmooth github page
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/cancertx.rda?raw=true"))
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/sm_predict_ex1.rda?raw=true"))

ggplot(sm_predict_ex1, aes(x = marker, y = .fitted)) +
  geom_line() +
  geom_ribbon(aes(ymin = .fitted.ll, ymax = .fitted.ul), alpha = 0.3) +
  geom_line(
    data = cancertx %>% filter(marker >= 5 & marker <= 15),
    aes(x = marker, y = survt1_marker),
    linetype = "dashed"
  )
```


## Example 2 - Multivariable

```{r remedy002}
library(plotly)
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/sm_predict_ex2.rda?raw=true"))

age.seq =
  sm_predict_ex2 %>%
  select(age) %>%
  distinct() %>%
  arrange(age) %>%
  pull()

marker.seq =
  sm_predict_ex2 %>%
  select(marker) %>%
  distinct() %>%
  arrange(marker) %>%
  pull()

# adding true risk to dataset
sm_predict_ex2 = 
  sm_predict_ex2 %>%
  mutate(
    survt1 = exp(-exp((1/50) * age - 0.2 * marker)),
    surv_diff = survt1 - .fitted
  )

# plot histogram of differences between fitted and true
sm_predict_ex2 %>%
  ggplot(aes(x = surv_diff)) +
  geom_histogram() +
  labs(
    title = "Distribution",
    x = "Difference in estimated survival and true survival"
  )

sm_predict_ex2 %>%
  select(marker, age, survt1) %>%
  tidyr::spread(age, survt1) %>%
  select(-marker) %>%
  plot_ly(z = ~ as.matrix(.),
          x = marker.seq,
          y = age.seq) %>%
  add_surface(showscale=FALSE) %>%
  layout(
    title = "True Risk of Death within 1 Year",
    scene = list(
      xaxis = list(title = "Marker"),
      yaxis = list(title = "Age"),
      zaxis = list(title = "Risk")
    ))

sm_predict_ex2 %>%
  select(marker, age, .fitted) %>%
  tidyr::spread(age, .fitted) %>%
  select(-marker) %>%
  plot_ly(z = ~ as.matrix(.),
          x = marker.seq,
          y = age.seq) %>%
  add_surface(showscale=FALSE) %>%
  layout(
    title = "Estimated Risk of Death within 1 Year",
    scene = list(
      xaxis = list(title = "Marker"),
      yaxis = list(title = "Age"),
      zaxis = list(title = "Risk")
    ))
```
