---
title: "sm_predict"
author: "Daniel D. Sjoberg"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{sm_predict}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Introduction

The `sm_predict()` function calculates kernel-smoothed predictions from regression models (i.e. outputs from the `predict()` function).  The following examples focus on time-to-event endpoints.  The example data set is simulated from a Cox regression model.

$$
h(t) = h_0(t)e^{\textbf{X}\beta}
$$

with $h_0(t) = 1$ and $\textbf{X}\beta = 0.02 * age - 0.2 * marker$.  The independent variables age and marker are independent from one another.  The example data set containts the following columns: 

    time    Years from cancer treatment to death
    age     Age at treatment
    marker  Marker level at treatment
    survt1  True 1 year survival probability

```{r dta, message=FALSE}
library(sjosmooth)
library(dplyr)
# loading data from sjosmooth github page
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/cancertx.rda?raw=true"))

cancertx %>% select(time, age, marker, survt1) %>% head() %>% knitr::kable()
```

The simualted data contains `r nrow(cancertx) %>% format(big.mark = ",", scientific=FALSE)` observations.

## Example 1 - Univariate

Kernel smoothing can be computationally intense on large data sets.  The following code was run to get the example object `sm_regression_ex1` and the results saved to  [GitHub](https://github.com/ddsjoberg/sjosmooth/tree/master/examples).

```
library(survival)
library(sjosmooth)
sm_predict_ex1 =
  sm_predict(
    data = cancertx,
    method = "coxph",
    formula = Surv(time) ~ marker,
    newdata = dplyr::data_frame(marker = seq(5, 15, 0.5), time = 1),
    type = "survival",
    verbose = TRUE
  )
```

The data in the `cancertx` data frame was simulated from a Cox regression model with linear predictor.  

$$
\textbf{X}\beta = 0.02 * age - 0.2 * marker
$$

In the model age and marker level are independent, and therefore, the univariate model the slope coefficient for marker remains $\beta_{marker} = -0.2$ 

```{r remedy001}
library(ggplot2)
# loading data and example results from sjosmooth github page
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/sm_predict_ex1.rda?raw=true"))

ggplot(sm_predict_ex1, aes(x = marker, y = .fitted)) +
  geom_line() +
  geom_ribbon(aes(ymin = .fitted.ll, ymax = .fitted.ul), alpha = 0.3) +
  geom_line(
    data = cancertx %>% filter(marker >= 5 & marker <= 15),
    aes(x = marker, y = survt1_marker),
    linetype = "dashed"
  ) +
  labs(
    y = "1 year survival probability"
  )
```

The dashed line is the true one year survival, and the solid line is the estimated survival from `sm_predict()`.

## Example 2 - Multivariable

In this example, the 1 year survival will be estimated by both age and marker level.  The following code was run in advance, and the results saved to [GitHub](https://github.com/ddsjoberg/sjosmooth/tree/master/examples).

```
sm_predict_ex2 <-
  sjosmooth::sm_predict(
    data = cancertx,
    method = "coxph",
    formula = Surv(time) ~ marker + age,
    newdata =
      list(
        marker = seq(7, 13, 1),
        age = seq(20, 40, 2)
      ) %>%
        purrr::cross_df() %>%
        dplyr::mutate(time = 1),
    type = "survival",
    verbose = TRUE
  )
```

```{r remedy002}
# loading example results
load(url("https://github.com/ddsjoberg/sjosmooth/blob/master/examples/sm_predict_ex2.rda?raw=true"))

# adding true risk to dataset, and calculating difference between estiamted and true rate
sm_predict_ex2 = 
  sm_predict_ex2 %>%
  mutate(
    survt1 = exp(-exp((1/50) * age - 0.2 * marker)),
    surv_diff = survt1 - .fitted
  )

# plot histogram of differences between fitted and true
sm_predict_ex2 %>%
  ggplot(aes(x = surv_diff)) +
  geom_histogram(bins = 25) +
  labs(
    title = "Difference Distribution",
    x = "Difference in estimated and true survival"
  )


```
 
The code below will create a 3-D surface plot.  The resulting HTML-widgets figures are too large to include in the package, however.

```{r}
# library(plotly)
# # extracting unique values of independent variables
# age.seq =
#   sm_predict_ex2 %>%
#   select(age) %>%
#   distinct() %>%
#   arrange(age) %>%
#   pull()
# 
# marker.seq =
#   sm_predict_ex2 %>%
#   select(marker) %>%
#   distinct() %>%
#   arrange(marker) %>%
#   pull()
# 
# # Plotting the true survival probabilities
# sm_predict_ex2 %>%
#   select(marker, age, survt1) %>%
#   tidyr::spread(age, survt1) %>%
#   select(-marker) %>%
#   plot_ly(z = ~ as.matrix(.),
#           x = marker.seq,
#           y = age.seq) %>%
#   add_surface(showscale=FALSE) %>%
#   layout(
#     title = "True Risk of Death within 1 Year",
#     scene = list(
#       xaxis = list(title = "Marker"),
#       yaxis = list(title = "Age"),
#       zaxis = list(title = "Survival Prob.")
#     ))
# 
# # Plotting the estimated survival probabilities from sm_predict()
# sm_predict_ex2 %>%
#   select(marker, age, .fitted) %>%
#   tidyr::spread(age, .fitted) %>%
#   select(-marker) %>%
#   plot_ly(z = ~ as.matrix(.),
#           x = marker.seq,
#           y = age.seq) %>%
#   add_surface(showscale=FALSE) %>%
#   layout(
#     title = "Estimated Risk of Death within 1 Year",
#     scene = list(
#       xaxis = list(title = "Marker"),
#       yaxis = list(title = "Age"),
#       zaxis = list(title = "Survival Prob.")
#     ))
```
